\chapter{Conclusion and Future Work}
\label{chap:conclusion}


This chapter completes this thesis by presenting a conclusion on the work and proposing future work.

\subsection*{Multi-Resolution Point Cloud Storage}

This thesis presents an out-of-core octree data structure that stores the point-cloud data in a cache file on the hard disc. While the leaf nodes of the octree contain the actual point-cloud data, the intermediate nodes store subsamples from its child nodes, thus creating a multi-resolution representation of the point cloud. The octree is paired with a level-of-detail decision that decides which part of the point cloud and resolution is rendered. This approach results in higher memory usage. However, the low-resolution representations of the point cloud allow for fast rendering and coarse but fast processing. 


\subsection*{User-Guided Shape Detection and Clustering}

This thesis shows an alternative use of the shape-detection algorithm by Schnabel et al. \cite{schnabel-2007-efficient} that lets the user control the regions that are segmented. By pointing to a region with the mouse, the system selects the most suitable octree node to be segmented. The octree's split decision ensures a consistent number of points per shape, so that the shape detection delivers results in interactive time. The shape detection is performed on the highest visible level of detail. Hence, this approach creates a multi-scale representation of the shapes by performing shape detection on nodes with different level of detail. To account for different point distributions in octree nodes, a dynamic $\epsilon$ threshold, based on the node's point density, is proposed for the shape detection. 

\par

As the size of a single shape is limited to the extents of the octree node it was detected in, this thesis proposes a shape clustering algorithm that determines if two shapes are matching and creates a larger coherent shape from multiple matching primitive shapes of different levels of detail that is then used to assist user interactions. 


\subsection*{Improvement of Interaction Techniques}

Several improvements to well-known two-dimensional interaction techniques are discussed in this thesis. We propose a way of pre-filtering points for interactions such that only points are considered that are approximated by a support shape, picked by the user. Classic point picking is improved so that only points are picked that belong to the support shape. Only using points that belong to a shape is especially useful when trying to pick points that are otherwise occluded or lie on the edge of a structure. Multiple ways exist to select regions in a point cloud. We utilize a support shape to improve the lasso selection. Classic lasso selection selects all points, whose projection lies inside the lasso on the near plane. The user selects `through` the point cloud, whereas when using a support shape, the user only selects points that lie on this support shape. The volumetric brush allows the user to select points that are in the foreground. Instead of consulting the depth buffer to retrieve the brush's position, the position of the cursor on the support shape is used. Thus, the brush does not follow the structures that are in the foreground but the curvature of the support shape and points can be selected that were occluded otherwise. 


\subsection*{Novel Interaction to Magnify Local Details}

The final task of this thesis was to design a novel interaction technique called \textit{shape-assisted local level-of-detail increment}. The level-of-detail rendering of the point cloud displays only a subset of points that can be rendered by the graphics card in real time. Thus, details may get culled and are not rendered. The interaction technique collects points from nodes that are not rendered due to their level-of-detail. These additional points are approximated by the support shape. Along the support shape, extra points are rendered that would otherwise be culled, therefore, allowing the user to get a more detailed look at structures. 


\section{Future Work}

As the focus of this thesis was the design of the user-guided shape detection and the assisted interactions, future work will focus on performance and robustness improvements on various fronts. The usage of a selection data structure, such as a selection octree \cite{scheiblauer2011out}, can improve the performance when selecting or editing the point cloud. Also, different ways can be explored to improve computation speed by taking advantage of the parallel architecture of the graphics card. 

\par

As the selection processing is performed asynchronously, feedback is not displayed immediately, but with a delay. To overcome this delay between interaction and receiving the result, a visual selection can be performed on the GPU \cite{rainer2016visual} to bridge this gap. This technique utilizes volumetric shadows to create a visual selection using the GPU's stencil buffer. 

\par

Shape detection is performed using an external library by Schnabel et al. \cite{schnabel-2007-software}. Multiple problems occur, such as non-termination or non-plausible shape matching. More work must be carried out to limit or eliminate these problems. The robustness of the shape detection, especially when detecting non-planar primitive shapes, suffers due to weak constraints. Alternatively, the shape detection can be implemented in such a way, that particular types of shapes are prioritized to reduce the amount of non-plausible shapes. 

\par

In the current application, edges between nodes with different levels of detail are visible. Regions with sparse points contain visible gaps. Adaptive point sizes \cite{scheiblauer-thesis} could be used to reduce the visual artifacts that come with this type of level-of-detail rendering. The individual point size for each octree node is controlled by the number of points and the weighted level of detail. Points in octree nodes with fewer points but larger extent have a larger size. 

\par 

The current point-cloud rendering system draws points multiple times. The overdraw of redundant points accounts for $\sim$12\% of the entire point budget. Future work on this domain will aim at the reduction of this overdraw. An intermediate node needs to be drawn only if the level-of-detail decision culls one of its children. If this is the case, the parent node contains non-duplicate point information that is not displayed by one of its children.