\section {Out-of-core Point-Clouds}
\label{sec:related_work_point_clouds}

As the size of modern point clouds often exceeds the available memory, specialized data structures and rendering techniques are needed that can handle such amounts of data. 

The QSplat system \cite{rusinkiewicz2000qsplat} was one of the earliest systems that were capable of handling datasets with well over hundred million points. It uses a hierarchy of bounding spheres that makes it easy to perform visibility culling and level-of-detail control. Each node only contains information on the bounding sphere, not the points itself. Leaf nodes represent a single point sample. Hence, the bounding sphere is the point itself. A node is rendered if it is a leaf or the benefit of traversing to the children is too low; otherwise, the children are traversed. If a node is rendered, a spherical splat is drawn as the node's bounding sphere. 

\par

Gobbetti and Marton \cite{gobbetti2004layered} propose a multi-resolution approach for rendering massive point clouds, called Layered Point Clouds (LPC). The point cloud is stored in a binary tree in chunks of approximately the same size. The multi-resolution model contains the same points as the point cloud, but grouped into chunks and organized in a level-of-detail representation. The root of the tree contains a subset of uniformly distributed samples of the point cloud. The remaining points are divided among the two subtrees that are further partitioned until the number of points is below a threshold value. Compared to the QSplat system, LPC reduces the cost of traversal on the CPU-side significantly and benefits from the parallel architecture of the GPU. This approach hides out-of-core latency by speculatively fetching data. 

\par

Dachsbauer et al. \cite{dachsbacher2003sequential} introduced Sequential Point Trees (SPT), a data structure that allows adaptive rendering of point clouds completely on the graphics card. A hierarchical point tree is not suited for fast vertex array-based sequential processing by the GPU. Therefore, the point tree's nodes are rearranged into a list, sorted by depth. Rendering then only needs to draw the first $p$ points, where $p$ is controlled by a level-of-detail decision. While their technique allows for adaptive rendering purely on the GPU, it only allows for a single level of detail at a time and can only contain a limited number of points. 

\par

Wimmer and Scheiblauer \cite{wimmer2006instant} introduce nested octrees to create a multi-resolution model similar to LPCs. A nested octree consists of an outer octree that is used for visibility culling and a memory-optimized SPT as inner octree for efficient point rendering. Each node contains a subsample of points from the point cloud. The union of the points from level 0 to a level $k$ creates the level-of-detail representation of the point cloud for level $k$. 

\par

Besides rendering and storing, modifications are a key task for out-of-core point cloud systems. Wand et al. \cite{wand2007interactive} describe an out-of-core octree that also stores a multi-resolution model of the point cloud. The basic idea is to store downsampled point clouds in the inner nodes where each inner node either contains randomly chosen or averaged points from its children. Hence, this octree stores additional points per level of detail. Points are inserted bottom-up, and the multi-resolution model is updated on the way up. Removing points works similarly. The point is deleted from the leaf first, and the ancestor nodes are updated afterward. 

\par

An improvement to nested octrees is the Modifiable Nested Octree (MNO) by Scheiblauer and Wimmer \cite{scheiblauer2011out}. The MNO enhances the nested octree by improving the performance of insertion and deletion operations. A grid replaces the SPTs from the inner nodes. When inserting a new point, an empty grid cell that contains the point is searched. When removing a point from the hierarchy, holes can occur. Therefore, points from a leaf node that intersects the grid cell of the removed point are pulled up. MNO is paired with an out-of-core point-cloud editing system. Since selecting points in an out-of-core setting is not trivial, the authors propose a structure, the so-called Selection Octree to store the selected points separately. This structure is a tool that allows to interactively change the visualization model without actually having to modify the original model permanently. 

\par

Until now, the discussed approaches focused on storage and rendering of point clouds. Wenzel et al. \cite{wenzel2014out} propose an out-of-core octree system that is tailored towards quick data updates and memory management. Parallel to the octree containing the point cloud, a cycle stack is used as a node history of in-memory data. This history allows to keep track of currently used nodes, prevents premature de-allocation of shared nodes, and manages the available memory. 


\section{Point-Based Rendering}
\label{sec:related_work_point_rendering}
Point clouds are discrete samplings of continuous surfaces and usually only provide position and color information. The lack of connectivity information disallows for polygon-based rasterization as used for triangle meshes. Pfister et al. \cite{pfister2000surfels} propose surface elements (surfels) to render complex geometric objects without connectivity information. A surfel is a point sampled from the surface of a geometric object and is rendered as an oriented disc. 

\par

Surface splatting by Zwicker et al. \cite{zwicker2001surface} is a technique for high-quality rendering of points with irregular spacing. An elliptical weighted filter is applied on a framebuffer that contains the points projected into screen space. Points closer to the center have a higher contribution to the pixel's color. Points with a depth difference above a particular threshold are excluded from the filtering so that only points that belong to the same surface contribute to the splat. This approach was developed in a software renderer. Bosch et al. \cite{botsch2005high} improved upon the original surface splatting algorithm by describing a GPU-driven approach to increase the performance. Their method consists of multiple render passes. First, a visibility pass determines which points are visible, then the attribute pass accumulates normal, and color attributes into a framebuffer, and finally, a normalization and shading pass normalizes the attributes and performs shading. 

\par

Most point-cloud datasets lack normal information, which is a requirement for drawing surface-aligned surfels for surface splatting. Instead, Scheiblauer \cite{scheiblauer2011consolidated} propose the usage of screen-aligned circles. To allow for higher-quality surface splatting, Preiner et al. \cite{preiner2012auto} describe a technique to calculate missing normals and radii in screen space during rendering. 

\par

More recently, Schütz and Wimmer \cite{schutz2015high} propose a nearest-neighbor-like interpolation technique to improve the visual quality of point-cloud renderings without the need of multiple render passes. Instead of an aligned disk, a screen-aligned quad is drawn per point, and the depth of each fragment is displaced to form the surface of a sphere, cone or paraboloid. The resulting images show strong similarities to a Voronoi diagram. Due to its simplicity and performance, this technique is used in this thesis as well to render the point cloud. 
